---
title: 'Tutorial 5: Co-occurrence analysis'
author: "Andreas Niekler, Gregor Wiedemann"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: yes
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_float: yes
highlight: tango
csl: springer.csl
bibliography: references.bib
---
```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```
This exercise will demonstrate how to perform co-occurrence analysis with **R** and the quanteda-package. It is shown how different significance measures can be used to extract semantic links between words. 

Change to your working directory, create a new R script, load the quanteda-package and define a few already known default variables.

```{r initalisierung, results='hide', message=FALSE, warning=FALSE}
options(stringsAsFactors = FALSE)
library(quanteda)

textdata <- read.csv("data/sotu.csv", sep = ";", encoding = "UTF-8")

sotu_corpus <- corpus(textdata$text, docnames = textdata$doc_id, docvars = data.frame(year = substr(textdata$date, 0, 4)))
```

# Sentence detection

The separation of the text into semantic analysis units is important for co-occurrence analysis. Context windows can be for instance documents, paragraphs or sentences or neighboring words. One of the most frequently used context window is the sentence. 

Documents are decomposed into sentences. Sentences are defined as a separate (quasi-)documents in a new corpus object of the quanteda-package. The further application of the quanteda-package functions remains the same. In contrast to previous exercises, however, we now use sentences which are stored as individual documents in the body.

Important: The sentence segmentation must take place *before* the other preprocessing steps because the sentence-segmentation-model relies on intact word forms and punctuation marks.

The following code uses a quanteda function to **reshape** the corpus into sentences. 


```{r warning=F, message=F}
# original corpus length and its first document
ndoc(sotu_corpus)

substr(texts(sotu_corpus)[1], 0, 200)

corpus_sentences <- corpus_reshape(sotu_corpus, to = "sentences")

ndoc(corpus_sentences)

texts(corpus_sentences)[1]
texts(corpus_sentences)[2]

```

CAUTION: The newly decomposed corpus has now reached a considerable size of `r ndoc(corpus_sentences)` sentences. Older computers may get in trouble because of insufficient memory during this preprocessing step.

Now we are returning to our usual pre-processing chain and apply it on the separated sentences.

```{r preprocessCorpus, cache=TRUE}
# Build a dictionary of lemmas
lemma_data <- read.csv("resources/baseform_en.tsv", encoding = "UTF-8")

# read an extended stop word list
stopwords_extended <- readLines("resources/stopwords_en.txt", encoding = "UTF-8")

# Preprocessing of the corpus of sentences
corpus_tokens <- corpus_sentences %>% 
  tokens(remove_punct = TRUE, remove_numbers = TRUE, remove_symbols = TRUE) %>% 
  tokens_tolower() %>% 
  tokens_replace(lemma_data$inflected_form, lemma_data$lemma, valuetype = "fixed") %>% 
  tokens_remove(pattern = stopwords_extended, padding = T)

# calculate multi-word unit candidates
sotu_collocations <- textstat_collocations(corpus_tokens, min_count = 25)
sotu_collocations <- sotu_collocations[1:250, ]

corpus_tokens <- tokens_compound(corpus_tokens, sotu_collocations)
```

Again, we create a document-term-matrix. Only word forms which occur at least 10 times should be taken into account. An upper limit is not set (`Inf` = infinite).

Additionally, we are interested in the joint occurrence of words in a sentence. For this, we do not need the exact count of how often the terms occur, but only the information whether they occur together or not. This can be encoded in a binary document-term-matrix. The parameter `weighting` in the control options calls the ` weightBin` function. This writes a 1 into the DTM if the term is contained in a sentence and 0 if not.

```{r binDTM, cache=TRUE}
minimumFrequency <- 10

# Create DTM, prune vocabulary and set binary values for presence/absence of types
binDTM <- corpus_tokens %>% 
  tokens_remove("") %>%
  dfm() %>% 
  dfm_trim(min_docfreq = minimumFrequency, max_docfreq = Inf) %>% 
  dfm_weight("boolean")
```

# Counting co-occurrences

The counting of the joint word occurrence is easily possible via a matrix multiplication (https://en.wikipedia.org/wiki/Matrix_multiplication) on the binary DTM. For this purpose, the transposed matrix (dimensions: nTypes x nDocs) is multiplied by the original matrix (nDocs x nTypes), which as a result encodes a term-term matrix (dimensions: nTypes x nTypes).

```{r message=FALSE}
# Matrix multiplication for cooccurrence counts
coocCounts <- t(binDTM) %*% binDTM
```

Let's look at a snippet of the result. The matrix has `nTerms` rows and columns and is symmetric. Each cell contains the number of joint occurrences. In the diagonal, the frequencies of single occurrences of each term are encoded.

```{r}
as.matrix(coocCounts[202:205, 202:205])
```

Interprete as follows: `r colnames(coocCounts)[204]` appears together `r coocCounts[204, 205]` times with `r colnames(coocCounts)[205]` in the `r nrow(binDTM)` sentences of the SUTO addresses. `r colnames(coocCounts)[204]` alone occurs `r coocCounts[204, 204]` times.

# Statistical significance

In order to not only count joint occurrence we have to determine their significance. Different significance-measures can be used. We need also various counts to calculate the significance of the joint occurrence of a term `i` (`coocTerm`) with any other term `j`:
* k - Number of all context units in the corpus
* ki - Number of occurrences of `coocTerm`
* kj - Number of occurrences of comparison term j
* kij - Number of joint occurrences of `coocTerm` and j

These quantities can be calculated for any term `coocTerm` as follows:

```{r}
coocTerm <- "spain"
k <- nrow(binDTM)
ki <- sum(binDTM[, coocTerm])
kj <- colSums(binDTM)
names(kj) <- colnames(binDTM)
kij <- coocCounts[coocTerm, ]
```

An implementation in *R* for Mutual Information, Dice, and Log-Likelihood may look like this. At the end of each formula, the result is sorted so that the most significant co-occurrences are at the first ranks of the list.

```{r results='hide', message=FALSE, warning=FALSE}
########## MI: log(k*kij / (ki * kj) ########
mutualInformationSig <- log(k * kij / (ki * kj))
mutualInformationSig <- mutualInformationSig[order(mutualInformationSig, decreasing = TRUE)]

########## DICE: 2 X&Y / X + Y ##############
dicesig <- 2 * kij / (ki + kj)
dicesig <- dicesig[order(dicesig, decreasing=TRUE)]

########## Log Likelihood ###################
logsig <- 2 * ((k * log(k)) - (ki * log(ki)) - (kj * log(kj)) + (kij * log(kij)) 
               + (k - ki - kj + kij) * log(k - ki - kj + kij) 
               + (ki - kij) * log(ki - kij) + (kj - kij) * log(kj - kij) 
               - (k - ki) * log(k - ki) - (k - kj) * log(k - kj))
logsig <- logsig[order(logsig, decreasing=T)]
```

The result of the four variants for the statistical extraction of co-occurrence terms is shown in a data frame below. It can be seen that frequency is a bad indicator of meaning constitution. Mutual information emphasizes rather rare events in the data. Dice and Log-likelihood yield very well interpretable contexts.

```{r}
# Put all significance statistics in one Data-Frame
resultOverView <- data.frame(
  names(sort(kij, decreasing=T)[1:10]), sort(kij, decreasing=T)[1:10],
  names(mutualInformationSig[1:10]), mutualInformationSig[1:10], 
  names(dicesig[1:10]), dicesig[1:10], 
  names(logsig[1:10]), logsig[1:10],
  row.names = NULL)
colnames(resultOverView) <- c("Freq-terms", "Freq", "MI-terms", "MI", "Dice-Terms", "Dice", "LL-Terms", "LL")
print(resultOverView)
```


# Visualization of co-occurrence

In the following, we create a network visualization of significant co-occurrences. 

For this, we provide the calculation of the co-occurrence significance measures, which we have just introduced, as single function in the file `calculateCoocStatistics.R`.  This function can be imported into the current R-Session with the `source` command.


```{r}
# Read in the source code for the co-occurrence calculation
source("calculateCoocStatistics.R")
# Definition of a parameter for the representation of the co-occurrences of a concept
numberOfCoocs <- 15
# Determination of the term of which co-competitors are to be measured.
coocTerm <- "california"
```

We use the imported function `calculateCoocStatistics` to calculate the co-occurrences for the target term *"california"*.

```{r}
coocs <- calculateCoocStatistics(coocTerm, binDTM, measure="LOGLIK")
# Display the numberOfCoocs main terms
print(coocs[1:numberOfCoocs])
```

To acquire an extended semantic environment of the target term, 'secondary co-occurrence' terms can be computed for each co-occurrence term of the target term. This results in a graph that can be visualized with special layout algorithms (e.g. Force Directed Graph). 

Network graphs can be evaluated and visualized in R with the `igraph`-package. Any graph object can be created from a three-column data-frame. Each row in that data-frame is a triple. Each triple encodes an edge-information of two nodes (source, sink) and an edge-weight value. 

For a term co-occurrence network, each triple consists of the target word, a co-occurring word and the significance of their joint occurrence. We denote the values with *from, to, sig*. 

```{r warning=FALSE}
resultGraph <- data.frame(from = character(), to = character(), sig = numeric(0))
```

The process of gathering the network for the target term runs in two steps. First, we obtain all significant co-occurrence terms for the target term. Second, we obtain all co-occurrences of the co-occurrence terms from step one.

Intermediate results for each term are stored as temporary triples named `tmpGraph`. With the `rbind` command ("row bind", used for concatenation of data-frames) all `tmpGraph` are appended to the complete network object stored in `resultGraph`.

```{r}
# The structure of the temporary graph object is equal to that of the resultGraph
tmpGraph <- data.frame(from = character(), to = character(), sig = numeric(0))

# Fill the data.frame to produce the correct number of lines
tmpGraph[1:numberOfCoocs, 3] <- coocs[1:numberOfCoocs]
# Entry of the search word into the first column in all lines
tmpGraph[, 1] <- coocTerm
# Entry of the co-occurrences into the second column of the respective line
tmpGraph[, 2] <- names(coocs)[1:numberOfCoocs]
# Set the significances
tmpGraph[, 3] <- coocs[1:numberOfCoocs]

# Attach the triples to resultGraph
resultGraph <- rbind(resultGraph, tmpGraph)

# Iteration over the most significant numberOfCoocs co-occurrences of the search term
for (i in 1:numberOfCoocs){
  
  # Calling up the co-occurrence calculation for term i from the search words co-occurrences
  newCoocTerm <- names(coocs)[i]
  coocs2 <- calculateCoocStatistics(newCoocTerm, binDTM, measure="LOGLIK")
  
  #print the co-occurrences
  coocs2[1:10]
  
  # Structure of the temporary graph object
  tmpGraph <- data.frame(from = character(), to = character(), sig = numeric(0))
  tmpGraph[1:numberOfCoocs, 3] <- coocs2[1:numberOfCoocs]
  tmpGraph[, 1] <- newCoocTerm
  tmpGraph[, 2] <- names(coocs2)[1:numberOfCoocs]
  tmpGraph[, 3] <- coocs2[1:numberOfCoocs]
  
  #Append the result to the result graph
  resultGraph <- rbind(resultGraph, tmpGraph[2:length(tmpGraph[, 1]), ])
}
```

As a result, `resultGraph` now contains all `numberOfCoocs * numberOfCoocs` edges of a term co-occurrence network.

```{r}
# Sample of some examples from resultGraph
resultGraph[sample(nrow(resultGraph), 6), ]
```

The package iGraph offers multiple graph visualizations for graph objects. Graph objects can be created from triple lists, such as those we just generated. In the next step we load the package iGraph and create a visualization of all nodes and edges from the object `resultGraph`.

```{r fig.width=10, fig.height=8, message=FALSE}
require(igraph)

# set seed for graph plot
set.seed(1)

# Create the graph object as undirected graph
graphNetwork <- graph.data.frame(resultGraph, directed = F)

# Identification of all nodes with less than 2 edges
verticesToRemove <- V(graphNetwork)[degree(graphNetwork) < 2]
# These edges are removed from the graph
graphNetwork <- delete.vertices(graphNetwork, verticesToRemove) 

# Assign colors to nodes (search term blue, others orange)
V(graphNetwork)$color <- ifelse(V(graphNetwork)$name == coocTerm, 'cornflowerblue', 'orange') 

# Set edge colors
E(graphNetwork)$color <- adjustcolor("DarkGray", alpha.f = .5)
# scale significance between 1 and 10 for edge width
E(graphNetwork)$width <- scales::rescale(E(graphNetwork)$sig, to = c(1, 10))

# Set edges with radius
E(graphNetwork)$curved <- 0.15 
# Size the nodes by their degree of networking (scaled between 5 and 15)
V(graphNetwork)$size <- scales::rescale(log(degree(graphNetwork)), to = c(5, 15))

# Define the frame and spacing for the plot
par(mai=c(0,0,1,0)) 

# Final Plot
plot(
  graphNetwork,				
  layout = layout.fruchterman.reingold,	# Force Directed Layout 
  main = paste(coocTerm, ' Graph'),
  vertex.label.family = "sans",
  vertex.label.cex = 0.8,
  vertex.shape = "circle",
  vertex.label.dist = 0.5,			# Labels of the nodes moved slightly
  vertex.frame.color = adjustcolor("darkgray", alpha.f = .5),
  vertex.label.color = 'black',		# Color of node names
  vertex.label.font = 2,			# Font of node names
  vertex.label = V(graphNetwork)$name,		# node names
  vertex.label.cex = 1 # font size of node names
)
```

# Optional exercises

1. Create term networks for "civil", "germany", "tax"
2. For visualization, at one point we filter for all nodes with less than 2 edges. By this, the network plot gets less dense, but we loose also a lot of co-occurring terms connected only to one term. Re-draw the network without this filtering. 
```{r optionalEx2, eval=T,cache=T,message=F,warning=F,echo=F,fig.width=10,fig.height=8}

# set seed for graph plot
set.seed(1)

# Create the graph object as undirected graph
graphNetwork <- graph.data.frame(resultGraph, directed = F)

# Assign colors to nodes (search term blue, others orange)
V(graphNetwork)$color <- ifelse(V(graphNetwork)$name == coocTerm, 'cornflowerblue', 'orange') 

# Set edge colors
E(graphNetwork)$color <- adjustcolor("DarkGray", alpha.f = .5)
# scale significance between 1 and 10 for edge width
E(graphNetwork)$width <- scales::rescale(E(graphNetwork)$sig, to = c(1, 10))

# Set edges with radius
E(graphNetwork)$curved <- 0.15 
# Size the nodes by their degree of networking (scaled between 5 and 15)
V(graphNetwork)$size <- scales::rescale(log(degree(graphNetwork)), to = c(5, 15))

# Define the frame and spacing for the plot
par(mai=c(0,0,1,0)) 

# Final Plot
plot(
  graphNetwork,				
  layout = layout.fruchterman.reingold,	# Force Directed Layout 
  main = paste(coocTerm, ' Graph'),
  vertex.label.family = "sans",
  vertex.label.cex = 0.8,
  vertex.shape = "circle",
  vertex.label.dist = 0.5,			# Labels of the nodes moved slightly
  vertex.frame.color = adjustcolor("darkgray", alpha.f = .5),
  vertex.label.color = 'black',		# Color of node names
  vertex.label.font = 2,			# Font of node names
  vertex.label = V(graphNetwork)$name,		# node names
  vertex.label.cex = 1 # font size of node names 
)
```
The plot may get very messy. Try lower values for `numberOfCoocs` to create a less dense network plot.


3. Separate the DTM into two time periods (year < 1968; year > = 1968). Represent the graphs for the term "family" for both time periods. Hint: Define functions for the sub processes of creating a binary DTM from a corpus object (`get_binDTM <- function(mycorpus)`) and for visualizing a co-occurrence network (`vis_cooc_network <- function(binDTM, coocTerm)`).
```{r optionalEx, eval=T,cache=T,message=F,warning=F,echo=F,fig.width=10,fig.height=8}
vis_cooc_network <- function(binDTM, coocTerm, graph_title) {
  
  require(igraph)
  
  # set seed for graph plot
  set.seed(1)
  
  coocs <- calculateCoocStatistics(coocTerm, binDTM, measure="LOGLIK")
  
  resultGraph <- data.frame(from = character(), to = character(), sig = numeric(0))
  # The structure of the temporary graph object is equal to that of the resultGraph
  tmpGraph <- data.frame(from = character(), to = character(), sig = numeric(0))
  
  # Fill the data.frame to produce the correct number of lines
  tmpGraph[1:numberOfCoocs, 3] <- coocs[1:numberOfCoocs]
  # Entry of the search word into the first column in all lines
  tmpGraph[, 1] <- coocTerm
  # Entry of the co-occurrences into the second column of the respective line
  tmpGraph[, 2] <- names(coocs)[1:numberOfCoocs]
  # Set the significances
  tmpGraph[, 3] <- coocs[1:numberOfCoocs]
  
  # Attach the triples to resultGraph
  resultGraph <- rbind(resultGraph, tmpGraph)
  
  # Iteration over the most significant numberOfCoocs co-occurrences of the search term
  for (i in 1:numberOfCoocs) {
    
    # Calling up the co-occurrence calculation for term i from the search words co-occurrences
    newCoocTerm <- names(coocs)[i]
    coocs2 <- calculateCoocStatistics(newCoocTerm, binDTM, measure="LOGLIK")
    
    #print the co-occurrences
    coocs2[1:10]
    
    # Structure of the temporary graph object
    tmpGraph <- data.frame(from = character(), to = character(), sig = numeric(0))
    tmpGraph[1:numberOfCoocs, 3] <- coocs2[1:numberOfCoocs]
    tmpGraph[, 1] <- newCoocTerm
    tmpGraph[, 2] <- names(coocs2)[1:numberOfCoocs]
    tmpGraph[, 3] <- coocs2[1:numberOfCoocs]
    
    #Append the result to the result graph
    resultGraph <- rbind(resultGraph, tmpGraph[2:length(tmpGraph[, 1]), ])
  }
  
  # Create the graph object as undirected graph
  graphNetwork <- graph.data.frame(resultGraph, directed = F)
  
  # Identification of all nodes with less than 2 edges
  verticesToRemove <- V(graphNetwork)[degree(graphNetwork) < 2]
  # These edges are removed from the graph
  graphNetwork <- delete.vertices(graphNetwork, verticesToRemove) 
  
  # Assign colors to nodes (search term blue, others orange)
  V(graphNetwork)$color <- ifelse(V(graphNetwork)$name == coocTerm, 'cornflowerblue', 'orange') 
  
  # Set edge colors
  E(graphNetwork)$color <- adjustcolor("DarkGray", alpha.f = .5)
  # scale significance between 1 and 10 for edge width
  E(graphNetwork)$width <- scales::rescale(E(graphNetwork)$sig, to = c(1, 10))
  
  # Set edges with radius
  E(graphNetwork)$curved <- 0.15 
  # Size the nodes by their degree of networking (scaled between 5 and 15)
  V(graphNetwork)$size <- scales::rescale(log(degree(graphNetwork)), to = c(5, 15))
  
  # Define the frame and spacing for the plot
  par(mai=c(0,0,1,0)) 
  
  # Final Plot
  plot(
    graphNetwork,				
    layout = layout.fruchterman.reingold,	# Force Directed Layout 
    main = graph_title,
    vertex.label.family = "sans",
    vertex.label.cex = 0.8,
    vertex.shape = "circle",
    vertex.label.dist = 0.5,			# Labels of the nodes moved slightly
    vertex.frame.color = adjustcolor("darkgray", alpha.f = .5),
    vertex.label.color = 'black',		# Color of node names
    vertex.label.font = 2,			# Font of node names
    vertex.label = V(graphNetwork)$name,		# node names
    vertex.label.cex = 1 # font size of node names 
  )
}

numberOfCoocs <- 15
minimumFrequency <- 10

binDTM1 <- binDTM[docvars(corpus_sentences)$year < "1968", ]
binDTM2 <- binDTM[docvars(corpus_sentences)$year >= "1968", ]

source("calculateCoocStatistics.R")
vis_cooc_network(binDTM1, "family", "Family (year < 1968)")
vis_cooc_network(binDTM2, "family", "Family (year >= 1968)")
```
