Tutorial 2: Processing of textual data
======================================

In this tutorial, we demonstrate how to read text data in R, tokenize texts and create a document-term matrix.

1. Reading CSV data into a corpus,
2. Create a document-term matrix,
3. Investigate Zipf's law on distribution of words.

Reading text data
-----------------

Set global options at the beginning.

```{r}
# Global options
options(stringsAsFactors = FALSE)
```

The `read.csv` command reads a **CSV** (Comma Separated Value) file from disk. Such files represent a table whose rows are represented by single lines in the files and columns are marked by a *separator* character within lines. Arguments of the command can be set to specify whether the CSV file contains a line with column names (header = `TRUE` or` FALSE`) and the character set.

We read a CSV containing 231 "State of the Union" addresses of the presidents of the United States. The texts are freely available from http://stateoftheunion.onetwothree.net. Our CSV file has the format: `"id";"speech_type";"president";"date";"text"`. Text is encapsualted into quotes (`"`). Since sepration is marked by `;` instead of `,`, we need to specify the separator char. 

```{r results='hide', message=FALSE, warning=FALSE}
# read csv into a data.frame
textdata <- read.csv("data/sotu.csv", header = TRUE, sep = ";", encoding = "UTF-8")
```

The texts are now available in a data frame together with some metadata (ID, speech type, president). Let us first see how many documents and metadata we have read.

```{r message=FALSE, warning=FALSE}
# dimensions of the data frame
dim(textdata)
# column names of text and metadata
colnames(textdata)
```

**How many speeches do we have per president?** This can easily be counted with the command `table`, which can be used to create a cross table of different values. If we apply it to a column, e.g. *president* of our data frame, we get the counts of the unique *president* values.

```{r eval=F, echo=T}
table(textdata[, "president"])
```
```{r eval=TRUE, echo=FALSE}
head(table(textdata[, "president"]), 12)
```

Now we want to transfer the loaded text source into a corpus object of the `tm`-package. First we load the package.

```{r results='hide', message=FALSE, warning=FALSE}
require(tm)
```

Then, we crate with *readTabular* a mapping between column names in the data frame and placeholders in the tm corpus object. A corpus object is created with the Corpus command. As parameter, the command gets the data source wrapped by a specific reader function (`DataframeSource`, other reader functions are available, e.g. for simple vectors). The reader control parameter takes the previously defined mapping of metadata as input.

```{r}
m <- list(ID = "id", content = "text")
myReader <- readTabular(mapping = m)
corpus <- Corpus(DataframeSource(textdata), readerControl = list(reader = myReader))
# have a look on the new corpus object
corpus
```

A corpus is an extension of R list objects. With the `[[]]` brackets, we can access single list elements, here documents, within a corpus.

```{r eval=F, echo=T}
# accessing a single document object
corpus[[1]]
# getting its text content
as.character(corpus[[1]])
```
```{r eval=T, echo=F}
corpus[[1]]
paste0(substring(as.character(corpus[[1]]), 0, 120), "...")
```

Success!!! We now have `r length(corpus)` speeches for further analysis available in a convenient tm corpus object!

Text statistics
---------------

A further aim of this exercise is to learn about statistical characteristics of text data. At the moment, our texts are represented as long character strings wrapped in document objects of a corpus. To analyze which word forms the texts contain, they must be __tokenized__. This means that all the words in the texts need to be identified and separated. Only in this way is it possible to count the frequency of individual word forms. A word form is also called **"type"**. The occurrence of a type in a text is a **"token"**.

For text mining, text are further transformed into a numeric representation. The basic idea is that the texts can be represented as statistics about the contained words (or other content fragments such as sequences of two words). The list of every distinct word form in the entire corpsu forms the **vocabulary** of a corpus. For each document, we can count how often each word of the vocabulary occurs in it. By this, we get a term frequency vector for each document. The dimensionality of this term vector corresponds to the size of the vocabulary. Hence, the word vectors have the same form for each document in a corpus. Consequently, multiple term vectors representing different documents can be combined into a matrix. This data structure is called **document-term matrix** (DTM).

The function `DocumentTermMatrix` of the tm package creates such a DTM. If this command is called without further parameters, the individual word forms are identified by using the "space" as the word separator.

```{r}
# Create a DTM (may take a while)
DTM <- DocumentTermMatrix(corpus)
# Show some information
DTM
# Dimensionality of the DTM
dim(DTM)
```

The dimensions of the DTM, `r nrow(DTM)` rows and `r ncol(DTM)` columns, match the number of documents in the corpus and the number of different word forms (types) of the vocabulary.

A first impression of text statistics we can get from a word list. Such a word list represents the frequency counts of all words in all documents. We can get that information easily from the DTM by summing all of its column vectors.

A so-called **sparse matrix** data structure is used for the document term matrix in the tm package (tm imports the `slam` package for sparse matrices). Since most entries in a document term vector are 0, it would be very inefficient to actually store all these values. A sparse data structure instead stores only those values of a vector/matrix different from zero. The *slam* package provides arithmetic operations on sparse DTMs.

```{r}
require(slam)
# sum columns for word counts
freqs <- col_sums(DTM)
# get vocabulary vector
words <- colnames(DTM)
# combine words and their frequencies in a data frame
wordlist <- data.frame(words, freqs)
# re-order the wordlist by decreasing frequency
wordIndexes <- order(wordlist[, "freqs"], decreasing = TRUE)
wordlist <- wordlist[wordIndexes, ]
# show the most frequent words
head(wordlist, 25)
```

The words in this sorted list have a ranking depending on the position in this list. If the word ranks are plotted on the x axis and all frequencies on the y axis, then the Zipf distribution is obtained. This is a typical property of language data and its distribution is similar for all languages.

```{r fig.width=8, fig.height=7}
plot(wordlist$freqs , type = "l", lwd=2, main = "Rank frequency Plot", xlab="Rank", ylab ="Frequency")
```

The distribution follows an extreme power law distribution (very few words occur very often, very many words occur very rare). The Zipf law says that the frequency of a word is reciprocal to its rank (1 / r). To make the plot more readable, the axes can be logarithmized.

```{r fig.width=8, fig.height=7}
plot(wordlist$freqs , type = "l", log="xy", lwd=2, main = "Rank-Frequency Plot", xlab="log-Rank", ylab ="log-Frequency")
```

In the plot, two extreme ranges can be determined. Words in ranks between ca. 10,000 and `r nrow(wordlist)` can be observed only 10 times or less. Words below rank 100 can be oberved more than 1000 times in the documents. The goal of text mining is to automatically find structures in documents. Both mentioned extreme ranges of the vocabulary often are not suitable for this. Words which occur rarely, on very few documents, and words which occur extremely often, in almost every document, do not contribute much to the meaning of a text. 

Hence, ignoring very rare / frequent words has many advantages:

* reducing the dimensionality of the vocabulary (saves memory)
* processing speed up
* better identification of meaningful structures.

To illustrate the range of ranks best to be used for analysis, we augment information in the rank frequency plot. First, we mark so-called **stop words**. These are words of a language that normally do not contribute to semantic information about a text. In addition, all words in the word list are identified which occur less than 10 times.

The `%in%` operator can be used to compare which elements of the first vector are contained in the second vector. At this point, we compare the words in the word list with a loaded stopword list (retrieved by the function `stopwords` of the tm package) . The result of the `%in%` operator is a boolean vector which contains TRUE or FALSE values.

A boolean value (or a vector of boolean values) can be inverted with the `!` operator (`TRUE` gets `FALSE` and vice versa). The `which` command returns the indices of entries in a boolean vector which contain the value `TRUE`.

We also compute indices of words, which occur less than 10 times. With a union set operation, we combine both index lists. With a setdiff operation, we reduce a vector of all indices (the sequence `1:nrow(wordlist)`) by removing the stopword indices and the low freuent word indices.

With the command "lines" the range of the remining indices can be drawn into the plot.

```{r fig.width=8, fig.height=7}
plot(wordlist$freqs, type = "l", log="xy",lwd=2, main = "Rank-Frequency plot", xlab="Rank", ylab = "Frequency")
englishStopwords <- stopwords("en")
stopwords_idx <- which(wordlist$words %in% englishStopwords)
low_frequent_idx <- which(wordlist$freqs < 10)
insignificant_idx <- union(stopwords_idx, low_frequent_idx)
meaningful_range_idx <- setdiff(1:nrow(wordlist), insignificant_idx)
lines(meaningful_range_idx, wordlist$freqs[meaningful_range_idx], col = "green", lwd=2, type="p", pch=20)
```

The green range marks the range of meaningful terms for the collection.

Optional exercises
-------------------------------------------------
1. Print out the word list without stop words and low frequent words.
```{r echo=FALSE}
head(wordlist[meaningful_range_idx, ], 25)
```
2. If you look at the result, are there any corpus specific terms that should also be considered as stop word?
3. What is the share of terms regarding the entire vocabulary which occurr only once in the corpus?
```{r echo=FALSE}
sum(wordlist$freqs == 1) / nrow(wordlist)
```
4. Compute the type-token ratio (TTR) of the corpus. the TTR is the fraction of the number of tokens divided by the number of types.
```{r echo=FALSE}
ncol(DTM) / sum(DTM)
```